Variational Monte Carlo on bosonic systems

### 2018-04-03 project_1/article/article.tex
Start on theory of VMC

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        v(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    oscillator trap.  In the latter potential the oscillatory frequency is
    different in the $z$-direction, i.e., $\omega_z \neq \omega$.  The full
    Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one-body operator, $h(\vf{r}_i)$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + v(\vf{r}_i),
    \end{align}
    for equal masses, and the two-body interaction operator, $w(\vf{r}_i,
    \vf{r}_j)$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
        \label{eq:two-body_interaction}
    \end{align}
    where $a$ is the hard sphere of the particle.  The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ is a variational parameter and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ is the \emph{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
        \label{eq:correlation_wavefunction}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$. In practice \autoref{eq:two-body_interaction} is
    unnecessary as the situtation where $r_{jk} \leq a$ automatically yields a
    probability density of zero from \autoref{eq:correlation_wavefunction} thus
    rejecting the state completely.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute force Metropolis-Hastings
        algorithm is that we might be spending much computational resources in
        an uninteresting part of configuration space. To make smarter moves we
        will use the Metropolis-Hastings algorithm with the additions of
        \emph{Importance Sampling}  (which will be discussed in due time). This
        alteration of the algorithm is dependent on the \emph{drift force} of
        the system\footnote{Also known as the \emph{quantum force}}.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$ as the Monte Carlo sampling involves
        choosing a single particle at a time.

@@

    \subsection{Interacting hard sphere bosons}
        Moving to the full elliptical system for $\beta \neq 1$ and setting $a
        \neq 0$ we can write the trial wavefunction as
        \begin{align}
            \Psi_T(\vf{r})
            &=
            \Phi_T(\vf{r})
            J(\vf{r}),
        \end{align}
        where $\Phi_T(\vf{r})$ is the same Slater permanent as in
        \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
        \emph{Jastrow factor} given by
        \begin{align}
            J(\vf{r})
            &=
            \exp\para{
                \sum_{j < l}^N u(r_{jl})
            },
        \end{align}
        where
        \begin{align}
            u(r_{jk})
            = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr]
            \equiv u_{jk}
        \end{align}
        The drift force for particle $k$ for this system becomes
        \begin{align}
            \vf{F}_k(\vf{r})
            &=
            2\Biggl(
                \frac{\nabla_k\phi_k}{\phi_k}
                + \sum_{m \neq k}^N
                \nabla_k u_{km}
            \Biggr).
        \end{align}

        The Laplacian of particle $k$ divided by the trial wavefunction is
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and Laplacian of the single
        particle functions yields
        \begin{gather}
            \frac{\nabla_k\phi_k}{\phi_k}
            =
            -2\alpha
            (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k),
            \\
            \frac{\nabla_k^2\phi_k}{\phi_k}
            =
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr).
        \end{gather}
        Repeating this for the correlation functions yields
        \begin{align}
            \dpd[]{u_{km}}{r_{km}}
            &=
            \frac{a}{r_{km}(r_{km} - a)},
            \\
            \dpd[2]{u_{km}}{r_{km}}
            &= \frac{a^2 - 2ar_{km}}{r_{km}^2(r_{km} - a)^2}.
        \end{align}
        We will avoid writing out the full expression for the local energy for
        this system due to the size of the equation.

@@

            Looking at the one-body part of the Hamiltonian we get
            \begin{align}
                H
                &=
                \sum_{k = 1}^N
                \brac{
                    -\frac{\hbar^2}{2m}\nabla_k^2
                    + \half m\brak{
                        \omega^2(x_k^2 + y_k^2)
                        + \omega_z^2 z_k^2
                    }
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also gets a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core sphere radius $a/a_{\text{ho}} =
            0.0043$.

>>>>

\section{Variational Monte Carlo}
    The variational principle states that for a given \emph{trial wavefunction},
    $\ket{\Psi_T}$, the expectation value of the Hamiltonian, $H$, will be an
    upper bound to the ground state energy, $E_0$ of the system.
    \begin{align}
        E_0 \leq E
        = \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}},
    \end{align}
    The complicated part of the variational principle is finding the minimum, or
    values close to the minimum. For a select few systems we can calculate this
    analytically, but for the interesting systems we quickly hit an intractable
    wall. To get around this we try to calculate the energy for a given
    Hamiltonian and a trial wavefunction for different variational parameters.
    The variance of the variational energy is given by
    \begin{align}
        \sigma^2
        &=
        \bra{\Psi_T}H^2\ket{\Psi_T}
        - \bra{\Psi_T}H\ket{\Psi_T}^2.
    \end{align}
    If we have found the ``true'' trial wavefunction $\ket{\Psi_T}$, i.e., the
    eigenfunction to the Hamiltonian $H$, we get $\sigma^2 = 0$. We are thus
    interested in trying to find the variational parameters in the trial
    wavefunction that minimizes the variance $\sigma^2$.

    \subsection{Monte Carlo integration}
        Unfortunately the evaluation of the expectation value of the energy for
        a given trial wavefunction involves a large integral.
        \begin{align}
            E
            &= \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
            =
            \frac{\int\dd\vf{r}\Psi_T^{*}(\vf{r})H\Psi_T(\vf{r})}
            {\int\dd\vf{r}|\Psi_T|^2},
        \end{align}
        where $\vf{r} = (\vf{r}_1, \vf{r}_2, \dots \vf{r}_N)$ are the
        positions\footnote{This can include both spatial and spin parts, but we
        will in the remainder of the project ignore spin.} of the particles
        contained in the system.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute force Metropolis-Hastings
        algorithm is that we might be spending much computational resources in
        an uninteresting part of configuration space. To make smarter moves we
        will use the Metropolis-Hastings algorithm with the additions of
        \emph{Importance Sampling}  (which will be discussed in due time). This
        alteration of the algorithm is dependent on the \emph{drift force} of
        the system\footnote{Also known as the \emph{quantum force}.}.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$ as the Monte Carlo sampling involves
        choosing a single particle at a time.

\section{Systems}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        v(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    oscillator trap.  In the latter potential the oscillatory frequency is
    different in the $z$-direction, i.e., $\omega_z \neq \omega$.  The full
    Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one-body operator, $h(\vf{r}_i)$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + v(\vf{r}_i),
    \end{align}
    for equal masses, and the two-body interaction operator, $w(\vf{r}_i,
    \vf{r}_j)$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
        \label{eq:two-body_interaction}
    \end{align}
    where $a$ is the hard sphere of the particle.  The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ is a variational parameter and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ is the \emph{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
        \label{eq:correlation_wavefunction}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$. In practice \autoref{eq:two-body_interaction} is
    unnecessary as the situtation where $r_{jk} \leq a$ automatically yields a
    probability density of zero from \autoref{eq:correlation_wavefunction} thus
    rejecting the state completely.

@@

    \subsection{Interacting hard sphere bosons}
        Moving to the full elliptical system for $\beta \neq 1$ and setting $a
        \neq 0$ we can write the trial wavefunction as
        \begin{align}
            \Psi_T(\vf{r})
            &=
            \Phi_T(\vf{r})
            J(\vf{r}),
        \end{align}
        where $\Phi_T(\vf{r})$ is the same Slater permanent as in
        \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
        \emph{Jastrow factor} given by
        \begin{align}
            J(\vf{r})
            &=
            \exp\para{
                \sum_{j < l}^N u(r_{jl})
            },
        \end{align}
        where
        \begin{align}
            u(r_{jk})
            = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr]
            \equiv u_{jk}
        \end{align}
        The drift force for particle $k$ in this system becomes
        \begin{align}
            \vf{F}_k(\vf{r})
            &=
            2\Biggl(
                \frac{\nabla_k\phi_k}{\phi_k}
                + \sum_{m \neq k}^N
                \nabla_k u_{km}
            \Biggr).
        \end{align}

        The Laplacian of particle $k$ divided by the trial wavefunction is
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
                \label{eq:laplacian_full_system}
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and Laplacian of the single
        particle functions yields
        \begin{gather}
            \frac{\nabla_k\phi_k}{\phi_k}
            =
            -2\alpha
            (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k),
            \\
            \frac{\nabla_k^2\phi_k}{\phi_k}
            =
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr).
        \end{gather}
        Repeating this for the correlation functions yields
        \begin{align}
            \dpd[]{u_{km}}{r_{km}}
            &=
            \frac{a}{r_{km}(r_{km} - a)},
            \\
            \dpd[2]{u_{km}}{r_{km}}
            &= \frac{a^2 - 2ar_{km}}{r_{km}^2(r_{km} - a)^2}.
        \end{align}
        The local energy is thus
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                + v(\vf{r}_k)
            },
        \end{align}
        where the Laplcian is given by \autoref{eq:laplacian_full_system}.  We
        will avoid writing out the full expression for the local energy for this
        system due to the size of the equation.

@@

            Looking at the one-body part of the Hamiltonian we get
            \begin{align}
                H
                &=
                \sum_{k = 1}^N
                \brac{
                    -\frac{\hbar^2}{2m}\nabla_k^2
                    + \half m\brak{
                        \omega^2(x_k^2 + y_k^2)
                        + \omega_z^2 z_k^2
                    }
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.
